---
title: "Introduction to cori.data.hu"
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.hu}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval       = FALSE,
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

`cori.data.hu` provides annual housing unit estimates from the U.S. Census Bureau's National, State, and County Housing Unit Totals, processed and stored in CORI's S3 data lake. Data cover 2000–present and measure the total housing stock — occupied and vacant units combined — built from the decennial Census baseline plus annual permitted construction, non-permitted construction, mobile home shipments, and estimated housing loss.

**Source:** U.S. Census Bureau, Population Estimates Program
**Coverage:** 2000–present, updated annually (release typically May)
**Geography:** County (5-digit FIPS), state (2-digit FIPS), national (`"00"`)

## Variables

```{r codebook, eval=TRUE}
library(cori.data.hu)

get_hu_codebook() |>
  dplyr::select(variable, label, unit, notes) |>
  knitr::kable()
```

## Reading data

All data are returned in long format: one row per `geoid / year / variable`.

```{r read}
df <- get_housing_units(geography = "county")

dplyr::glimpse(df)
```

Filter to specific variables, years, or counties:

```{r read-filtered}
# Housing units per 1,000 population for New Hampshire counties
nh_hu <- get_housing_units(
  variables = "housing_units_per_1k_people",
  geoids    = grep("^33", unique(df$geoid), value = TRUE),
  years     = 2010:2024
)

dplyr::glimpse(nh_hu)
```

## Rural vs. Nonrural

Housing stock per capita diverges between rural and nonrural counties. The chart below shows population-weighted average housing units per 1,000 population by rural status using the CBSA 2023 rural definition.

```{r rural-nonrural}
library(cori.charts)
library(ggplot2)
library(ruraldefinitions)
library(dplyr)

load_fonts()

rural_avg <- df |>
  filter(variable == "housing_units_per_1k_people") |>
  left_join(select(cbsa_2023, geoid, is_rural), by = "geoid") |>
  filter(!is.na(is_rural), !is.na(value), !is.na(agg_var)) |>
  group_by(year, is_rural) |>
  summarise(
    value = sum(value * agg_var, na.rm = TRUE) / sum(agg_var, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(rural_avg, aes(x = year, y = value, color = is_rural)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("Rural" = "#2F6E9B", "Nonrural" = "#7EBDC2"),
    labels = c("Rural" = "Rural counties", "Nonrural" = "Nonrural counties")
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Rural counties carry more housing units per capita than nonrural",
    subtitle = "Population-weighted average housing units per 1,000 population, 2000\u20132024",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    caption  = "Source: CORI analysis of U.S. Census Bureau Housing Unit Estimates."
  )
```

## County Spotlight: Grafton County, NH

```{r grafton}
grafton <- get_housing_units(
  variables = "housing_units_per_1k_people",
  geoids    = c("33009", "33", "00")
) |>
  mutate(
    group = case_when(
      geoid == "33009" ~ "Grafton County, NH",
      geoid == "33"    ~ "New Hampshire",
      geoid == "00"    ~ "United States"
    ),
    group = factor(group, levels = c("Grafton County, NH", "New Hampshire", "United States"))
  )

ggplot(grafton, aes(x = year, y = value, color = group, linetype = group, linewidth = group)) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  scale_color_manual(values = c(
    "Grafton County, NH" = "#2F6E9B",
    "New Hampshire"      = "#7EBDC2",
    "United States"      = "#9DA7B0"
  )) +
  scale_linetype_manual(values = c(
    "Grafton County, NH" = "solid",
    "New Hampshire"      = "dashed",
    "United States"      = "dotted"
  )) +
  scale_linewidth_manual(values = c(
    "Grafton County, NH" = 1.2,
    "New Hampshire"      = 0.8,
    "United States"      = 0.8
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Housing units per 1,000 residents — Grafton County, NH",
    subtitle = "Compared to New Hampshire and the United States, 2000\u20132024",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    linetype = NULL,
    linewidth = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Housing Unit Estimates.\n",
      "Population denominator from Census Population Estimates Program via cori.data.pep."
    )
  )
```
