---
title: "Introduction to cori.data.hu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.hu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

## What Is This Data?

The U.S. Census Bureau's **National, State, and County Housing Unit Totals**
are annual estimates of the total number of housing units in existence —
occupied and vacant combined. Published under the Population Estimates Program,
these estimates measure the total housing stock, built from the 2020 Census
baseline plus annual permitted construction, non-permitted construction, mobile
home shipments, and estimated housing loss.

**Source:** U.S. Census Bureau, Population Estimates Program
**Coverage:** 2000–present, updated annually (release typically May)
**Geography:** County (5-digit FIPS), state (2-digit FIPS), national (`"00"`)

## Variable Reference

```{r codebook}
library(cori.data.hu)

get_hu_codebook() |> knitr::kable()
```

## Quick Start

```{r read}
library(dplyr)

df <- read_hu_from_s3()

dplyr::glimpse(df)
```

## Rural vs. Nonrural

Housing stock per capita diverges between rural and nonrural counties. The
chart below shows population-weighted average `housing_units_per_1k_people`
by rural status using the CBSA 2023 rural definition.

```{r rural-nonrural}
library(cori.charts)
library(ggplot2)
library(ruraldefinitions)

load_fonts()

rural_avg <- df |>
  filter(variable == "housing_units_per_1k_people", nchar(geoid) == 5) |>
  left_join(select(cbsa_2023, geoid, is_rural), by = "geoid") |>
  filter(!is.na(is_rural), !is.na(value), !is.na(agg_var)) |>
  group_by(year, is_rural) |>
  summarise(
    value = sum(value * agg_var, na.rm = TRUE) / sum(agg_var, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(rural_avg, aes(x = year, y = value, color = is_rural)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("Rural" = "#2F6E9B", "Nonrural" = "#7EBDC2"),
    labels = c("Rural" = "Rural counties", "Nonrural" = "Nonrural counties")
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2024, by = 5),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.1),
    expand = expansion(mult = c(0.02, 0.1))
  ) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Rural counties carry more housing units per capita than nonrural",
    subtitle = "Housing units per 1,000 population, 2000\u20132024",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    caption  = "Source: CORI analysis of U.S. Census Bureau Housing Unit Estimates."
  )
```

## County Spotlight: Grafton County, NH

```{r grafton}
grafton <- df |>
  filter(
    variable == "housing_units_per_1k_people",
    geoid %in% c("33009", "33", "00")
  ) |>
  mutate(
    group = case_when(
      geoid == "33009" ~ "Grafton County, NH",
      geoid == "33"    ~ "New Hampshire",
      geoid == "00"    ~ "United States"
    ),
    group = factor(group, levels = c("Grafton County, NH", "New Hampshire", "United States"))
  )

grafton_latest <- grafton |>
  filter(geoid == "33009", year == max(year)) |>
  pull(value)

grafton_year <- grafton |>
  filter(geoid == "33009") |>
  filter(year == max(year)) |>
  pull(year)

ggplot(grafton, aes(x = year, y = value, color = group,
                    linetype = group, linewidth = group)) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  scale_color_manual(values = c(
    "Grafton County, NH" = "#2F6E9B",
    "New Hampshire"      = "#7EBDC2",
    "United States"      = "#9DA7B0"
  )) +
  scale_linetype_manual(values = c(
    "Grafton County, NH" = "solid",
    "New Hampshire"      = "dashed",
    "United States"      = "dotted"
  )) +
  scale_linewidth_manual(values = c(
    "Grafton County, NH" = 1.2,
    "New Hampshire"      = 0.8,
    "United States"      = 0.8
  )) +
  scale_x_continuous(
    breaks = seq(2000, 2024, by = 5),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.1),
    expand = expansion(mult = c(0.02, 0.1))
  ) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = paste0(
      round(grafton_latest, 1),
      " housing units per 1,000 residents in Grafton County"
    ),
    subtitle = paste0("Housing units per 1,000 population, 2000\u2013", grafton_year),
    x        = NULL,
    y        = NULL,
    color    = NULL,
    linetype = NULL,
    linewidth = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Housing Unit Estimates.\n",
      "Population denominator from Census Population Estimates Program via cori.data.pep."
    )
  )
```


